New results in $\sfrac{k}{n}$ Power-Hours

### 2013-12-29 powerhour/paper/paper.tex
First submission

\usepackage{url}
\usepackage{code}
% \usepackage{cite}
% \usepackage{eclbkbox}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{code}
% \usepackage{amscd}
% \usepackage{xy}
\usepackage{graphicx}
% \usepackage{fancyhdr}
% \usepackage{color}
% \usepackage[dark,all,bottom,landscape,timestamp]{draftcopy}
% \usepackage{everypage}

% \pagestyle{empty}

@@

\usepackage{natbib}
% \usepackage{inlinebib}

@@

\maketitle

\begin{abstract}
Something about the k-n hours

\end{abstract}

@@

In order to efficiently enumerate BM5L, I improved the algorithm
again. Observe that the ``drink'' action associated with a rule
usually does not affect anything but the final outcome. The only
exception is that in the rule for \fullcup, the player must drink if
passing in a state other than \fullcup. Putting this aside for a
moment, note that we can just count the number of times each rule was
executed for each player, producing an $s$-dimensional vector $[ d_1,
  \ldots d_s ]$ for each player. That player is able to achieve many
different drink totals, specifically, $d_1 \times r_1 + \ldots + d_s
\times r_s$ where $r_i$ is 1 if the player should drink on that rule
and 0 if not. Simulating a game this way is even more like abstract
interpretation (we leave a concrete value free and compute a formula
rather than an integer), and allows us to evaluate many concrete games
at once. At the end, we simply plug in every legal value for $r_i$ for
each player and insert those games into the database. This last step
is where we must tend to the exception around \fullcup. We may not set
$r_0$ to 0 if the player ever passes \fullcup\ in a non-full state. A
very close approximation would be to insist that $r_0 = 1$ if the rule
in the \fullcup\ position does not output as \fullcup, but this is
inexact, as that rule may never be executed\footnote{We may be able to
  argue that in that case, there always exists another game that does
  not violate this condition. But I think it is simpler to just
  implement the rules.} Instead, during simulation we keep track of
whether each player ever actually passed a non-full cup from the
\fullcup\ state. If so, then we force $r_0 = 1$, which attends to this
special case. \label{sec:linear}

@@

We already have exact results for two players in BM5L, so the next
uncharted territory is BM6L. The result after apparent convergence
appears in Figure~\ref{fig:samplebm6l}. The sampling procedure runs
for many hours before plateauing overnight with 95.54\% of the matrix
filled. This suggests that BM6L is not universal for two players, or
else the configurations for the missing cells are extremely rare.
% XXX never give up! It just jumped to 95.65.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{sample60min2players6states.png}
\end{center}\vspace{-0.1in}
\caption{ 25.6 billion samples of legal two-player BM6L
  configurations. Darker cells represent outcomes that occur more
  often; cells that are pure white never occurred and are likely to be
  unattainable. Note that the intensity represents the rank of
  occurrence, not the magnitude; in actuality, outcomes like $\langle
  0, 0 \rangle$ occur {\em much} more often than others.
  %
  95.54\% of the cells are filled.
  % XXX new version with more filled...!
}
\label{fig:samplebm6l}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{sample60min2players7states.png}
\end{center}\vspace{-0.1in}
\caption{ 22.1 billion samples of legal two-player BM7L
  configurations. Fewer configurations were sampled than in
  Figure~\ref{fig:samplebm6l} because they take somewhat longer than
  6-state games to simulate, and a smaller proportion of random games
  are legal.
  %
  All but four of the $(60 + 1)^2 = 3721$ cells are attainable: 
  Permutations of $\langle 11, 53 \rangle$ and $\langle 49, 53 \rangle$
  remain elusive. Note that 53 drinks was also unattainable in a solo
  BM7L power-hour; this may in some sense be the ``hardest'' number of
  shots to drink in \bmsl.
}
\label{fig:samplebm7l}
\end{figure}

@@

Note that all of the two-dimensional figures resemble one another even
though they are fundamentally different (adding players, adding
states, adding random trials). This suggests that the combinatorial
problem (``what outcomes are reachable from finite state machines that
look kind of like this?'') has some geometric structure.

Some of the patterns are easy to explain. The top-left half of the
matrix is more populous than the bottom right, for example. This is
because we can bound the total number of drinks by $60 / c$, where $c$
is the number of cups active in the game (same as the number of cups
in the starting states). The top-left half is the region where this
sum is less than or equal to 30; in two player games, both players
must start with a cup in order to get an outcome in the bottom-right
half. We also see distinct clumps around 0, 15, 30, 45, 60; divisors
of 60. This is intuitive because the expressive power of BMML comes
from the ability to form cycles of cup states and drink on some
fraction of them. Clumps are formed around these values because of the
possibility of preludes leading into the cycles
(Figure~\ref{fig:solo31}) that either drink (adding to the total) or
don't (subtracting from it). Minor clumps form as echoes between the
major ones, because a player may participate in two cycles of
different length (Figure~\ref{fig:dual45}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{powerday3.pdf}
\end{center}\vspace{-0.1in}
\caption{All possible outcomes for the first two players in 3-player
  power days. These are the same games as the 3-player power hours,
  but at this scale makes it clear the groupings and their sparsity in
  the limit. Lines plotted from $\langle 0,0 \rangle$ to $langle 60,
  k_2 \rangle$ and $\langle k_1, 60 \rangle$ show significant
  structure, but don't explain some of the interior points. These are
  probably games where a player participates in two cycles of
  different periods.
}
\label{fig:powerday3}
\end{figure}

\section{Known bounds}

Define \bmsl, where $s$ is the number of cup states. BMML=BM3L. Find
$s$ such that a solo or dual power hour achieves any $k$.

\begin{enumerate}
  \item With one player, we have exact bounds on what is possible in
    the generalized case. With 8 cup states, a single player can
    drink 0--60 shots. Since each player can just play independently,
    this result extends to any number of players in BM8L.
  \item With two players, we know that BM5L (XXX check) does not allow
    all combinations of $\langle k_1, k_2 \rangle$ shots, though BM5L
    can achieve any $\langle k_1, k_2 \rangle$ where $k_2 \in \{ 0, 1
    \}$. No other $k_2$ can be used universally, though of course many
    other combinations are possible
    (Figure~\ref{fig:composite2player}).
  \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
    unlikely, given that random exploration plateaus at about 95.54\%.
  \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
    possible, given how close BM7L is for a single player. We know it
    can get 99.89\%, at least!
  \item With three players one player can drink any number of shots in
    BMML with two other players helping.
\end{enumerate}

\section{Conjectures} \label{sec:conjectures}

With two friends, you can drink any amount

Teetotaller: When $\langle k_1, \ldots, k_i, \ldots, k_p \rangle$ is possible,
so is $\langle k_1, \ldots, 0, \ldots, k_p \rangle$ (under what conditions)?

\section{Some stuffff}

Hi

>>>>

\usepackage{url}
\usepackage{code}
% \usepackage{cite}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

\pagestyle{empty}

@@

\usepackage{natbib}

@@

\maketitle \thispagestyle{empty}

\begin{abstract}
We correct for inebriated missteps, using computational methods to
establish new bounds in generalized \kn Power-Hour theory.
\end{abstract}

@@

In order to efficiently enumerate BM5L, I improved the algorithm
again. Observe that the ``drink'' action associated with a rule
usually does not affect anything but the final outcome. The only
exception is that in the rule for \fullcup, the player must drink if
passing in a state other than \fullcup. Putting this aside for a
moment, note that we can just count the number of times each rule was
executed for each player, producing an $s$-dimensional vector $[ d_1,
  \ldots, d_s ]$ for each player. That player is able to achieve many
different drink totals, specifically, $d_1 \times r_1 + \ldots + d_s
\times r_s$ where $r_i$ is 1 if the player should drink on that rule
and 0 if not. Simulating a game this way is even more like abstract
interpretation (we leave a concrete value free and compute a formula
rather than an integer), and allows us to evaluate many concrete games
at once. At the end, we simply plug in every legal value for $r_i$ for
each player and insert those games into the database. This last step
is where we must tend to the exception around \fullcup. We may not set
$r_0$ to 0 if the player ever passes \fullcup\ in a non-full state. A
very close approximation would be to insist that $r_0 = 1$ if the rule
in the \fullcup\ position does not output as \fullcup, but this is
inexact, as that rule may never be executed.\!\footnote{We may be able to
  argue that in that case, there always exists another game that does
  not violate this condition. But I think it is simpler to just
  implement the rules.} Instead, during simulation we keep track of
whether each player ever actually passed a non-full cup from the
\fullcup\ state. If so, then we force $r_0 = 1$, which attends to this
special case. \label{sec:linear}

@@

We already have exact results for two players in BM5L, so the next
uncharted territory is BM6L. The result after apparent convergence
appears in Figure~\ref{fig:samplebm6l}. The sampling procedure runs
for many hours before plateauing overnight with 95.65\% of the matrix
filled. This suggests that BM6L is not universal for two players, or
else the configurations for the missing cells are extremely
rare.\!\footnote{This is definitely a possibility, as new cells were
still appearing after exploring tens of billions of samples. However,
the gap here seems quite large.}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{sample60min2players6states.png}
\end{center}\vspace{-0.1in}
\caption{ 37.1 billion samples of legal two-player BM6L
  configurations. Darker cells represent outcomes that occur more
  often; cells that are pure white never occurred and are likely to be
  unattainable. Note that the intensity represents the rank of
  occurrence, not the magnitude; in actuality, outcomes like $\langle
  0, 0 \rangle$ occur {\em much} more often than others.
  %
  95.65\% of the cells are filled.
}
\label{fig:samplebm6l}
\end{figure}

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=0.90 \linewidth]{sample60min2players7states.png}
\end{center}\vspace{-0.1in}
\caption{ 22.1 billion samples of legal two-player BM7L
  configurations. Fewer configurations were sampled than in
  Figure~\ref{fig:samplebm6l} because they take somewhat longer than
  6-state games to simulate, and a smaller proportion of random games
  are legal.
  %
  All but four of the $(60 + 1)^2 = 3721$ cells are attainable: 
  Permutations of $\langle 11, 53 \rangle$ and $\langle 49, 53 \rangle$
  remain elusive. Note that 53 drinks was also unattainable in a solo
  BM7L power-hour; this may in some sense be the ``hardest'' number of
  shots to drink in \bmsl.
}
\label{fig:samplebm7l}
\end{figure}

@@

Note that all of the two-dimensional figures resemble one another even
though they are fundamentally different (adding players, adding
states, adding random trials). Even samples from BMML with three
players (3D projected to 2D), which is shown in Figure~\ref{fig:3d},
produces a similar pattern. This suggests that the combinatorial
problem (``what outcomes are reachable from finite state machines that
look kind of like this?'') has some geometric structure.

Some of the patterns are easy to explain. The top-left half of the
matrix is more populous than the bottom right, for example. This is
because we can bound the total number of drinks by $60 / c$, where $c$
is the number of cups active in the game (same as the number of cups
in the starting states). The top-left half is the region where this
sum is less than or equal to 30; in two player games, both players
must start with a cup in order to get an outcome in the bottom-right
half. We also see distinct clumps around 0, 15, 30, 45, 60; these
correspond to simple fractions (``drink every other time''; ``drink
three of four times'') of 60. This is intuitive because the expressive
power of BMML comes from the ability to form cycles of cup states and
drink on some fraction of them. Clumps are formed around these values
because of the possibility of preludes leading into the cycles
(Figure~\ref{fig:solo31}) that either drink (adding to the total) or
don't (subtracting from it). Minor clumps form as echoes between the
major ones, because a player may participate in two cycles of
different length (Figure~\ref{fig:dual45}).

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{sample60min3players3states.png}
\end{center}\vspace{-0.1in}
\caption{490 million samples of three-player BMML Power Hours. The
  cube is rotated $15\,^{\circ}$ along each axis, the top-left corner is
  $\langle 0, 0, 0 \rangle$, and the bottom right is $\langle 60, 60,
  60 \rangle$. }
\label{fig:3d}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{powerday3.pdf}
\end{center}\vspace{-0.1in}
\caption{All possible outcomes for the first two players in 3-player
  power days. These are the same games as the 3-player power hours,
  but at this scale makes it clear the groupings and their sparsity in
  the limit. Lines plotted from $\langle 0,0 \rangle$ to $\langle 60,
  k_2 \rangle$ and $\langle k_1, 60 \rangle$ show significant
  structure, but don't explain some of the interior points. These are
  probably games where a player participates in two cycles of
  different periods.
}
\label{fig:powerday3}
\end{figure}

The less extreme \kn\ Power-Three-Hours appears in
Figure~\ref{fig:powerthreehours}.

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{sample180min2players7states.png}
\end{center}\vspace{-0.1in}
\caption{6,456,764,116 samples of two-player BM7L Power Three-Hours.
}
\label{fig:powerthreehours}
\end{figure}

\section{Conclusion}

This section summarizes the known bounds for \bmsl, and states some
conjectures, before concluding.

\subsection{Known results}
\begin{enumerate}
  \item With one player, we have exact bounds on what is possible in
    the generalized case. With 8 cup states, a single player can
    drink 0--60 shots. Since each player can just play independently,
    this result extends to any number of players in BM8L. With fewer
    than 8 cup states, not every $k$ can be achieved alone.
  \item With two players in BM3L or BM4L, it is not possible for one
    of the players to drink every $k$ even if the other player helps her
    out.
  \item With two players, we know that BM5L does allow one player to
    drink any $k_1$ if the other player assists. In fact we can
    achieve any $\langle k_1, k_2 \rangle$ where $k_2 \in \{ 0, 1 \}$.
    No other $k_2$ can be used universally, though of course many
    other combinations are possible
    (Figure~\ref{fig:composite2player}).
  \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
    unlikely, given that random exploration plateaus with about 95.65\%
    of the grid filled.
  \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
    possible, but there are four (really two with symmetry) unsolved
    combinations.
  \item With three players in BM3L, one player can drink any number of
    shots if the other two players help.
\end{enumerate}

\subsection{Conjectures} \label{sec:conjectures}

\paragraph{Freedom: With two friends, you can drink any amount.} We know
that in a three-player game of BM3L, one of the three players can
drink the $k$ of her choice. This straightforwardly extends to $3
\times p$-player BM3L games. The {\bf Freedom} conjecture is that
with $p + 2$ players in BM3L, $p$ of them may have their choice of
$k_1 \ldots k_n$ drinks. If this conjecture fails, it probably fails
for 4 players, which might have a feasible enumeration strategy.

\paragraph{Teetotaller: Someone can drink nothing.} When $\langle k_1, 
\ldots, k_i, \ldots, k_p \rangle$ is achievable in \bmsl, so is
$\langle k_1, \ldots, 0, \ldots, k_p \rangle$. This conjecture would
be trivial if not for the rule that requires us to drink the contents
of a full cup if we want to pass it in a different state. This
conjecture is true for all the graphics presented in this
paper;\footnote{Actually, it is not established for (only) 11 minutes
  in Figure~\ref{fig:powerthreehours}, but this is not a proper
  \bmsl\ game as it takes place over 180 minutes. There may very well
  be solutions for 11 anyway; this is just a sample.} we can see that
cells in the 0 column are always filled when some other cell in that
row is filled. If this conjecture fails, it probably fails for BM1L or
BM2L which have the least freedom per player.

\medskip

In this paper I present some new results in \kn\ Power-Hour theory, as
well as correct the historical record of some inebriated missteps. We
see that stochastic simulation, abstract interpration, and sampling
are powerful tools for solving combinatorial drinking problems. We
established some firm results for the classic game and some bounds for
generalizations, as well as informally looked at some visualizations
of its geometric structure. However, there are still several open
problems in this field that demand further study.
